Water desalination using polyelectrolyte hydrogel. Gibbs ensemble modelling

### 2022-07-01 gibbs.tex
results up to desalination

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
%\usepackage[utf8]{inputenc}
\usepackage{amssymb}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\newcommand{\ie}{\textit{i.~e.} }
\newcommand{\cref}{c^{\ominus}}
%\newcommand{\cs}{c_{s}}
\newcommand{\kT}{k_\mathrm{B}T}
\newcommand{\kB}{k_\mathrm{B}}
\newcommand{\lb}{l_\mathrm{B}}
\newcommand{\NA}{N_{\mathrm{A^-}}}
\newcommand{\muna}{\mu_\mathrm{Na^+}}

\newcommand{\mucl}{\mu_\mathrm{Cl^-}}
\newcommand{\muca}{\mu_\mathrm{Ca^{2+}}}
\newcommand{\muh}{\mu_\mathrm{H^+}}
\newcommand{\mua}{\mu_\mathrm{A^-}}
\newcommand{\muha}{\mu_\mathrm{HA}}
\newcommand{\muoh}{\mu_\mathrm{OH}}

\newcommand{\cna}{c_\mathrm{Na^+}}
\newcommand{\ccl}{c_\mathrm{Cl^-}}
\newcommand{\cca}{c_\mathrm{Ca^{2+}}}
\newcommand{\ch}{c_\mathrm{H^+}}
\newcommand{\cp}{c_\mathrm{p}}
\newcommand{\nna}{n_\mathrm{Na^+}}
\newcommand{\ncl}{n_\mathrm{Cl^-}}
\newcommand{\ncleq}{\widetilde{N}_\mathrm{Cl^-}}
\newcommand{\nca}{N_\mathrm{Ca^{2+}}}

\newcommand{\superin}{^\mathrm{in}}
\newcommand{\subin}{_\mathrm{in}}
\newcommand{\subi}{_\mathrm{i}}

\newcommand{\gel}{^\mathrm{gel}}
\newcommand{\tot}{^\mathrm{tot}}
\newcommand{\out}{^{out}}
\newcommand{\coh}{c_\mathrm{OH}}
\newcommand{\bulk}{^{\mathrm{b}}}
\renewcommand{\H}{\mathrm{H^+}}
\newcommand{\A}{\mathrm{A^-}}
\newcommand{\AH}{\mathrm{AH}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}
\newcommand{\todoi}[1]{\todo[inline]{#1}}
\setuptodonotes{fancyline, color=blue!30, size=\tiny}

\newcommand{\cl}{\mathrm{Cl^-}}
\newcommand{\br}{\mathrm{Br^-}}
\newcommand{\na}{\mathrm{Na^+}}
\newcommand{\h}{\mathrm{H^+}}
\newcommand{\ka}{\mathrm{K^+}}
\newcommand{\oh}{\mathrm{OH^-}}
\newcommand{\ca}{\mathrm{Ca^{2+}}}
\newcommand{\mg}{\mathrm{Mg^{2+}}}
\newcommand{\so}{\mathrm{SO_4^{2-}}}

\newcommand{\EI}{E_{\mathrm{I}}}
\newcommand{\EII}{E_{\mathrm{II}}}
\newcommand{\SI}{S_{\mathrm{I}}}
\newcommand{\SII}{S_{\mathrm{II}}}
\newcommand{\NI}{N_{\mathrm{I}}}
\newcommand{\NII}{N_{\mathrm{II}}}
\newcommand{\VI}{V_{\mathrm{I}}}
\newcommand{\VII}{V_{\mathrm{II}}}
\newcommand{\FI}{F_{\mathrm{I}}}
\newcommand{\FII}{F_{\mathrm{II}}}
\newcommand{\nnaI}{N^{\na}_{\mathrm{I}}}
\newcommand{\nclI}{N^{\cl}_{\mathrm{I}}}
\newcommand{\nnaII}{N^{\na}_{\mathrm{II}}}
\newcommand{\nclII}{N^{\cl}_{\mathrm{II}}}

\newcommand{\Ka}{K_{\mathrm{A}}}
\newcommand{\pKa}{\mathrm{p}\Ka}
\newcommand{\pK}{\mathrm{p}K}
\newcommand{\pH}{\mathrm{pH}}
\newcommand{\mol}{\mathrm{mol}}
\newcommand{\molperl}{\mathrm{mol/l}}
\newcommand{\kg}{\mathrm{kg}}
\newcommand{\res}{^{\mathrm{res}}}
\newcommand{\pHres}{\pH\res}
\newcommand{\pHgel}{\pH\gel}
\newcommand{\cs}{c_{\mathrm{s}}}
%\newcommand{\cs}{c_{\mathrm{salt}}}
\newcommand{\csres}{\cs\res}
%\newcommand{\Vgel}{V\gel}
\newcommand{\Vgel}{V_\mathrm{gel}}
\newcommand{\Ngel}{N_\mathrm{gel}}
\newcommand{\Vgeleq}{\widetilde{\Vgel}}
\newcommand{\Pgel}{P\gel}
\newcommand{\Pres}{P\res}
\newcommand{\Vout}{V_\mathrm{out}}
\newcommand{\Vbox}{V_\mathrm{box}}
\newcommand{\PE}{polyelectrolyte{}}

\newcommand{\reffig}[1]{Figure~\ref{#1}}
\newcommand{\refeq}[1]{Equation~\ref{#1}{}}
\usepackage{afterpage}
\newcommand\blankpage{%
    \null
    \thispagestyle{empty}%
    \addtocounter{page}{-1}%
    \newpage}

\newcommand{\mytitle}{Gibbs ensemble for \PE{} hydrogel}
\newcommand{\etal}{\textit{et al.}{}}

\@ifundefined{showcaptionsetup}{}{%
 \PassOptionsToPackage{caption=false}{subfig}}
\usepackage{subfig}
\makeatother

@@

\paragraph{Compression in open system.}

First we run a set of simulations modeling the gel compression in \emph{open system}, \ie in equilibrium with big bath of certain salinity, $\cs$. 
The simulations are run for a set of different simulation box volumes. 
Each simulation returns the averaged pressure, $P$, and the number of $\cl$ ions present in the simulation box, $\ncl$.

In order to get the partial pressure of the gel, we substitute from $P$ the osmotic pressure of ions in reservoir: $P^{gel}=P-P^{res}$.
$P^{res}$, in turn, we calculate by running a separate simulation of a reservoir, containing ionic gas in equilibrium with the bath of the same salinity $\cs$. 
The gel partial pressure, $\Pgel$, is the pressure which needs to be applied to the gel via solvent permeable filter to compress the gel to a molar specific volume, $V$.

The dependencies of $\Pgel$ versus gel molar volume $\Vgel$ (the volume calculated per one mol of gel segment) are present in Figure \ref{fig: PV} for a set of different salinity. 
These dependencies corresponding to \emph{open system} are represented here by solid lines. 
For example, blue solid line illustrates the compression (or swelling) of the gel in equilibrium with a reservoir of salinity, $\cs=0.063$ mol/l. 
The points where the pressure equals to zero, $\Pgel=0$, (indicated by filled circles) are the gel \emph{free swelling equilibrium} states. 
It is seen, that the increase of salinity shifts the free swelling equilibrium towards smaller volumes, \ie the gel shrinks with increasing salinity. 
In general, all the solid curves shift towards smaller volumes with increasing salinity. 
This effect is well known, and is typical for all branched \emph{strong} polyelectrolytes. 
It is caused by the decrease of ions osmotic pressure and by screening of electrostatic interactions \cite{Zhulina2000,Landsgesel2020a}.\footnote{The salinity dependence of the size of \emph{weak} polyelectrolyte gel is in general non-monotonic \cite{Rud2018} but the case of weak polyelectrolyte is beyond the scope of this study. } %

\paragraph{Compression in closed system.}

\begin{figure*}[t]
\subfloat[The gel partial pressure vs gel molar volume \label{fig: PV}]
{\includegraphics[width=0.48\textwidth]{figures/fig_PV_pub}}
\hspace{0.02\textwidth}
\subfloat[Supernate salinity vs gel molar volume\label{fig: CV}]
{\includegraphics[width=0.48\textwidth]{figures/fig_CV_pub}}
\caption{
The compression of the gel in open system (solid lines) and in closed system (dotted lines). 
Each solid curve corresponds to different salinity of the reservoir $\cs$ (see legend). Shadowed area limit the states with applied pressure below zero and above 5 bar.\label{fig: PV and CV}}
\todoi{replace $V$ by $\Vgel$ in Figures}
\end{figure*}

Each set of simulations in an \emph{open system} defines the free swelling equilibrium state of the gel for each salinity. 
Namely, each salinity correspond to a certain gel molar volume, and to a certain amount of ions in gel in equilibrium state, $\cs\longrightarrow$, $\left(\Vgeleq,\ \ncleq\right)$.
These equilibrium states we use as starting ones for the gel compression in \emph{closed system}. 
Thus, we simulate the compression of the gel in the closed volume, $\Vbox$, which is equal to $\Vgeleq$, and contains $\ncleq$ of ion pairs (apart from neutralizing counterions). 
So, we prepare two systems, one~--- simulating the gel of the volume $\Vgel$, and another~--- simulating the supernatant solution of the volume $\Vout = \Vbox - \Vgel$; in this way the total volume containing the gel and the supernatant solution remains constant.
Also, the ion pairs of the amount $\ncleq$ remain all the time constant and are shared between the two volumes. %For each different initial states we run a set of simulations emulating the compression of the gel, \ie varying simultaneously $\Vgel$ and $\Vout$. 
%At any all the ions are getting redistributed between the systems but their total amount remains constant and equal to $\ncl^{0}$.

The process of the gel compression in \emph{closed} system is depicted in Figure~\ref{fig: PV} (dotted lines). 
There the gel partial pressure is depicted as function of the gel molar volume.
In this plot, the blue dotted line, for example, illustrates the compression of the gel equilibrated with the solution of salinity $\cs=0.063$ mol/l, in a volume which gel has at zero pressure. 
Since, in this case, the volumes of the gel $\Vgel$ and of the supernatant solution $\Vout$ are comparable, the compression of the gel affects the salinity of supernatant solution. 
The salinity now decreases, changing from $\cs^{0}=0.063$ at free swelling equilibrium state to 0.045 mol/l at $P\gel=5$ bar.

The change of salinity is seen from Figure~\ref{fig: CV}. 
In this Figure the gel swelling/compression processes are illustrated in different coordinates, \ie salinity of supertnatant outside the gel versus the gel molar volume, $\cs(\Vgel)$.\todo{we should stick to only 'supernate' or 'supernatant'} 
%Similarly, as in Figure~\ref{fig:fig PV}, the solid lines correspond to compression in \emph{open} system and dotted lines~--- in \emph{closed} system. 
In this coordinates, all the open system compressions turn to the horizontal lines, which reflects the constant salinity, but the compressions in closed system indicate the decrease of $\cs$s.
% Maybe "In this coordinates, each open system compression turns to the horizontal line, which reflects the constant salinity, whereas each compression in closed system shows the decrease of $\cs$s." would be better

Although the salinity during compression in \emph{open} system remains constant, the number of ions in a compression volume, i.e. in the volume where the gel is compressed, changes.
As the compression volume $\Vbox$ we chose the volume of the free swelling equilibrium state of the gel, $\Vbox = \widetilde{\Vgel}$ 
In the Figure \ref{fig: NV} we depicted the number of $\cl$ ions in the volume $\Vbox$, calculated per unit volume, $\ncl/\Vbox$, as a function of the gel molar volume. 
The depicted value can be considered as the density of $\cl$ ions averaged out over the volume $\Vbox$.

As it is expected in case of closed system these dependencies look like horizontal lines, whereas in open system, the amount of ions in compression volume increases with compression. 
This implies, that the compression of the gel in open system pulls out the ions from the bath to the compression volume. 
And vice versa, the swelling of the gel pushes ions out to the bath.

Finally in Figure \ref{fig: CN} the same processes are depicted in coordinates number of particles in compression volume --- salinity of the outside solution. 
In this coordinates both ways of the compression, in \emph{open} and in \emph{closed} system, appear as straight vertical and horizontal lines correspondingly.

In our study we modeled the compression of the gel in equilibrium with reservoirs of 40 different salinities, ranging from 0.001 to 0.5 mol/l. 
Every open system compression resulted a defined free swelling equilibrium state, \ie the the gel molar volume and the number of ions in gel $\left(\Vgeleq,\ \ncleq\right)$ at zero pressure.
These states were used as the initial ones for the compression in closed systems. 
All the corresponding dependencies are depicted in Figures~\ref{fig: PV and CV} and \ref{fig: NV and CN} as thin grey dashed lines. 
The states corresponding to $P^{gel}=0$, 5 and 10 bar pressure are marked by open circles, squares and crosses respectively. 
The non-shadowed areas on the Figures limit the states corresponding to gel partial pressure ranging between 0 and 5 bar.

@@

\paragraph{Desalination scheme.}

As we have seen, the compression of the gel in a closed system affects the salinity, whereas the compression in an open system affects the amount of ions in a volume where the gel is compressed in. 
Here we show how to employ these phenomena for water desalination. 
The highlighted colored lines on plots of Figures \ref{fig: PV and CV} and \ref{fig: NV and CN} form a sequence of the following one by each other swellings and compressions of the hydrogel, correspondingly in \emph{open} and \emph{closed} systems. 
This sequence forms the water desalination process.
Starting from swelling the gel in the \emph{open} system at high salinity ($\cs=0.091$ mol/l, solid black line), the gel gets compressed in \emph{closed} system until the pressure reaches 5 bar (dashed black line). 
Then the same gel swells in a reservoir of a bit lower salinity (\ie $\cs=0.064$ mol/l, light blue line). 
After swelling, the gel is compressed again by the pressure of 5 bar in \emph{closed} system (dashed light blue line).  
Then it is put swelling in a reservoir of even smaller salinity ($\cs=0.045$ mol/l, solid yellow line). 
And so on. 
The chain of alternating swelling and compression steps ends up when salinity gets equal to $\cs=4\cdot10^{-3}$ mol/l after compression in closed system (dashed magenta line).

>>>>

\input{definitions.tex}
\makeatother

@@

\paragraph{Compression in open system.}
\begin{figure*}[t]
\subfloat[The gel partial pressure vs gel molar volume \label{fig: PV}]
{\includegraphics[width=0.48\textwidth]{figures/fig_PV_pub}}
\hspace{0.02\textwidth}
\subfloat[Supernate salinity vs gel molar volume\label{fig: CV}]
{\includegraphics[width=0.48\textwidth]{figures/fig_CV_pub}}
\caption{
The compression of the gel in open system (solid lines) and in closed system (dotted lines). 
Each solid curve corresponds to different salinity of the reservoir $\cs$ (see legend). Shadowed area limit the states with applied pressure below zero and above 5 bar.\label{fig: PV and CV}}
\todoi{replace $V$ by $\Vgel$ in Figures}
\end{figure*}
First, we run a set of simulations modeling the gel compression in \emph{open system}, \ie in equilibrium with a big bath of certain salinity, $\cs$. 
The simulations are run for a set of different simulation box volumes. 
Each simulation returns the averaged pressure, $P$, and the number of $\cl$ ions present in the simulation box, $\ncl$.

In order to get the partial pressure of the gel, we substitute from $P$ the osmotic pressure of ions in the reservoir: $P^{gel}=P - P^{res}$.
$P^{res}$, in turn, we calculate by running a separate simulation of a reservoir, containing ionic gas in equilibrium with the bath of the same salinity $\cs$. 
The gel partial pressure, $\Pgel$, is the pressure that needs to be applied to the gel via a solvent permeable filter to compress the gel to a molar-specific volume, $\Vgel$.

The dependencies of $\Pgel$ on the gel molar volume $\Vgel$ are present in Figure \ref{fig: PV} for a set of different salinities. 
These dependencies (the case of \emph{open system}) are represented by solid lines. 
For example, the blue solid line illustrates the compression (or swelling) of the gel in equilibrium with a reservoir of salinity, $\cs=0.063$ mol/l. 
The points where the pressure equals zero, $\Pgel=0$, (indicated by filled circles) are the gel \emph{free swelling equilibrium} states. 
It is seen, that the increase of salinity shifts the free swelling equilibrium towards smaller volumes, \ie the gel shrinks with increasing salinity. 
In general, all the solid curves shift towards smaller volumes with increasing salinity. 
This effect is well known and is typical for all branched \emph{strong} polyelectrolytes. 
It is caused by the decrease of ions osmotic pressure and by a screening of electrostatic interactions \cite{Zhulina2000, Landsgesel2020a}.\footnote{The salinity dependence of the size of \emph{weak} polyelectrolyte gel is in general non-monotonic. We discuss this case in \cite{Rud2018}. } %

@@

\paragraph{Compression in closed system.}
Each set of simulations in an \emph{open system} defines the free swelling equilibrium states of the gel for each salinity, that is the state where the gel partial pressure equals zero. 
These states are characterized by the corresponding molar volume of the gel, $\Vgel^0$ and the densities of ions in gel \{$\nna^0$, $\ncl^0$\} 
(Index ``0'' stands for zero bar applied pressure).
We use the free swelling equilibrium states as the starting ones for the gel compression in \emph{closed system}. 
Thus, we simulate the compression of the gel in the closed volume, $\Vbox$, which is equal to $\Vgel^0$, and contains $\ncl^0$ of $\cl$ ions and the $\na$ ions in the amount neutralizing the system,$\ncl^0$.

We prepare two systems, one~--- simulating the gel of the volume $\Vgel$, and another~--- simulating the supernatant solution of the volume $\Vout = \Vbox - \Vgel$.
Also, the ions of the amount $\Ncl^0$ and $\Nna^0$ are shared between the two volumes.

The process of the gel compression in the \emph{closed} system is depicted in Figure~\ref{fig: PV} (dotted lines) as the dependence of gel partial pressure versus the gel molar volume.
In this plot, the blue dotted line, for example, illustrates the compression of the gel equilibrated with the solution of salinity $\cs=0.063$ mol/l, in a volume which gel has at zero pressure. 
Since, in this case, the volumes of the gel $\Vgel$ and of the supernate $\Vout$ are comparable, the compression of the gel affects the salinity of supernate, $\cs$. 
The supernate salinity now decreases, changing from $\cs^{0}=0.063$ at the free swelling equilibrium state to $\cs^{5}=0.045$ mol/l in the state with applied pressure $P\gel=5$ bar (index ``5'' stands for 5 bar).

The \reffig{fig: CV} illustrates the same swelling/compression processes, but in different coordinates: \ie salinity of supernate versus the gel molar volume, $\cs(\Vgel)$.
In these coordinates, all the open system compressions show up as horizontal lines, which reflects the constant salinity, whereas the compressions in the closed system demonstrate the decrease of $\cs$.

Although the salinity during compression in \emph{open} system remains constant, the number of ions in a compression volume, i.e. in the volume where the gel is compressed (or swells), changes.
As the compression volume $\Vbox$ we chose the volume of the free swelling equilibrium state of the gel, $\Vbox = \Vgel^0$.
In the Figure \ref{fig: NV} we depicted the number of $\cl$ ions in the volume $\Vbox$, calculated per unit volume, $\ncl =$ $\Ncl/\Vbox$, as a function of the gel molar volume. 
The depicted value can be considered as the density of $\cl$ ions averaged out over the volume $\Vbox$.

As it is expected in case of closed system these dependencies look like horizontal lines, whereas in open system, the amount of ions in compression volume increases with compression. 
This implies, that the compression of the gel in open system pulls out the ions from the bath to the compression volume, $\Vbox$. 
And vice versa, the swelling of the gel pushes ions out to the bath.

Finally, in Figure \ref{fig: CN} the same processes are depicted in coordinates $\ncl$ --- $\cs$. 
In these coordinates, both ways of the compression, in \emph{open} and in \emph{closed} system, appear as straight vertical and horizontal lines correspondingly.

In our study, we modeled the compression of the gel in equilibrium with reservoirs of 40 different salinities, ranging from 0.001 to 0.5 mol/l. 
Every open system compression resulted in a defined free swelling equilibrium state, which we used as the initial ones for the compression in closed systems. 
All the corresponding dependencies are depicted in Figures~\ref{fig: PV and CV} and \ref{fig: NV and CN} as thin grey dashed lines (some of them are highlighted and colored).
The states corresponding to $P^{gel}=0$, 5 and 10 bar pressure are marked by open circles, squares, and crosses respectively. 
The non-shadowed areas in the Figures highlight the states in which the gel partial pressure ranges between 0 and 5 bar.

\paragraph{Desalination scheme.}
As we have seen, the compression of the gel in a closed system affects the salinity, whereas the compression in an open system affects the amount of ions in a volume where the gel is compressed in. 
Here we show how to employ these phenomena for water desalination. 
The highlighted colored lines on plots of Figures \ref{fig: PV and CV} and \ref{fig: NV and CN} form a sequence of the following one by each other swellings and compressions of the hydrogel, correspondingly in \emph{open} and \emph{closed} systems. 
This sequence forms the water desalination process.
Starting from swelling the gel in the \emph{open} system at high salinity ($\cs=0.091$ mol/l, solid black line), the gel gets compressed in \emph{closed} system until the pressure reaches 5 bar (dashed black line). 
Then the same gel swells in a reservoir of a bit lower salinity (\ie $\cs=0.064$ mol/l, light blue line). 
After swelling, the gel is compressed again by the pressure of 5 bar in \emph{closed} system (dashed light blue line).  
Then it is put swelling in a reservoir of even smaller salinity ($\cs=0.045$ mol/l, solid yellow line). 
And so on. 
The chain of alternating swelling and compression steps ends up when salinity gets equal to $\cs=4\cdot10^{-3}$ mol/l after compression in closed system (dashed magenta line).
